AGB Stars in the IR: Physical Characteristics, Galactic Spatial Distribution, and Chemical Impact of the Milky Way AGB Population

### 2015-07-28 paper1/paper1_data.tex
Changes made up to and including section 2.4

Our principal data source is the merged WISE-2MASS catalog. In order to derive
WISE-2MASS color-based selection criteria, we use several training samples
described below: a heterogeneous sample of bright \agb\ stars listed in 
SIMBAD database and more homogneous fainter samples selected using 
optical time-domain surveys OGLE-III and MACHO. The latter samples 
include \agb\ stars from the LMC and SMC and are also used to 
calibrate color-absolute magnitude relations that underlie our distance
estimates. In order to assess what infrared populations could contribute to
contamination of selected \agb\ candidates, we utilize a number of SDSS
catalogs, also described below. We conclude this section by describing
how these catalogs auxiliary catalogs were  positionally merged with the 
WISE-2MASS catalog and summarize the object counts before and after quality control in Table 1 {\color{red} [reference for table 1 when made]}.

\subsection{WISE-2MASS catalog}
Write a few sentences about the WISE survey...

\subsection{AGB stars from SIMBAD database}
The SIMBAD database\footnote{\url{http://simbad.u-strasbg.fr/simbad/}} is comprised of astrometry, classifications, and photometry for all types of astronomical objects from a variety of surveys.  Classifications are hierarchical, and are classified with increasing specificity based on the available data, though each star can have multiple classifications. AGB stars have subcategories of C-stars ([C/O] $>$ 1) and S-stars ([C/O] $\sim$ 1). Where subclassification is unclear, the main classification remains as AGB star. OH/IR stars are subsets of peculiar stars that specifically show strong OH maser emission and strong IR emission, identifying them clearly as AGB stars.  Mira-type variables are a subset of variable stars, and are classified based upon their periodicity. Semi-regular variables are not included as they contain an unknown mixture of AGB stars and red supergiants. The sample of AGB stars from SIMBAD was obtained by querying all objects with types classified as C-stars (23,628), S-stars (1,514), AGB stars (4,459), OH/IR stars (1,265), and Mira-type variables (10,828). Without duplicate objects between the five types, there is a total of 39,348 AGB stars from SIMBAD.

\subsection{AGB stars from optical time-domain surveys}
Many of the identifications in the SIMBAD catalog are from spectroscopic studies, however many more AGB stars have been found through their characteristic optical periodicity. The \emph{Optical Gravitational Lens Experiment} (OGLE) and the \emph{MAssive Compacy Halo Objects} (MACHO) survey are two time-domain microlensing surveys that have captured and catalogued such variability. In Sections~\ref{sec:macho} and \ref{sec:ogle} we disc.uss the origins of the data from each source and how they were then used for this study

\subsubsection{MACHO}\label{sec:macho}
The \emph{MACHO} project \citep{1992ASPC...34..193A,1997ApJ...482...89A} was a two-color optical microlensing survey searching for massive compact halo objects in the Milky Way bulge and the Magellanic Clouds. As a consequence of the time-domain survey, the \emph{MACHO} catalog contains 8 years of observations for several million stars, many of which were observed to be variable. 
Using Fourier decomposition, \cite{2008PhDT.........6F} identified and classified a population of 46,832 LPVs in the LMC, 3,943 in the SMC {\color{red}[check this]} , and {\color{red}[some other number]} in the Milky Way Bulge with matches to 2MASS. We note that due to high scatter and uncertainty of Galactic objects in period-magnitude space, we only take sequences 1 and 2 (Mira-type variables) for the Milky Way Bulge sample. Each object carries observations in Kron-Cousins $V$ \& $R$, and 2MASS photometry. AGB stars are separated from RGBs by identifying the tip of the RGB ($K_s < 12.3$ in the LMC and $K_s < 12.8$ in the SMC). {\color{red}[some small tidbit about the reduction procedure]}.

\subsubsection{OGLE-III}\label{sec:ogle}
The \emph{OGLE-III} Catalog of Variable Stars (CVS) \citep{2008AcA....58...69U,2009AcA....59..239S,2011AcA....61..217S} is a subset of the overall \emph{OGLE-III} experiment, containing roughly 10 years of observations in the $V$- and $I$-bands of over 120,000 variable stars. With high-precision photometry \citep[$~0.001$ mag][]{2007AcA....57..201S}, these observations saturate at $I = 12.5$ mag, and are limited at the faint end at $I = 20.5$ mag \citep{2001AcA....51..303Z}.

@@

\subsection{Imaging and spectroscopic catalogs from SDSS}
A few sentences of introduction...
\subsubsection{Galactic sources} 
\subsubsection{Extragalactic populations}

>>>>

Our principal data source is the merged WISE-2MASS catalog. In order to derive
WISE-2MASS color-based selection criteria, we use %several training samples
%described below: a heterogeneous sample of bright \agb\ stars listed in 
%SIMBAD database and more homogneous fainter samples selected using 
%optical time-domain surveys OGLE-III and MACHO. The latter samples 
%include \agb\ stars from the LMC and SMC and are also used to 
%calibrate color-absolute magnitude relations that underlie our distance
%estimates. In order to assess what infrared populations could contribute to
%contamination of selected \agb\ candidates, we utilize a number of SDSS
%catalogs, also described below. We conclude this section by describing
%how these catalogs auxiliary catalogs were  positionally merged with the 
%WISE-2MASS catalog and summarize the object counts before and after quality control in Table 1 
a homogenous sample of AGB stars from the OGLE-III Catalog of Variable Stars supplying objects in the Magellanic Clouds. This sample serves to calibrate color-color and color-absolute magnitude
relations that underlie our distance estimates. In order to assess what infrared populations
could contribute to contamination of selected \agb\ candidates, we utilize
extragalactic sources from SDSS data release 7 pulled from the NYU Value-added Galaxy Catalog,
the SDSS unofficial Luminous Red Galaxy sample,
and young stellar objects obtained identified in \wise\ data, also described below. 
We conclude this section by describing how these auxiliary catalogs were positionally merged
with the WISE-2MASS catalog, and summarize the object counts before and after quality control in 
Table 1 {\color{red} [reference for table 1 when made]}.

\subsection{WISE-2MASS catalog}
In this study, we rely heavily on data from the \allwise\, extension of the \wise\, survey, combining data from the initial All-Sky Data Release, the 3-band cryogenic data release, and the NEOWISE post-cryogenic data release \citep{2013wise.rept....1C}. The initial \wise\,All-Sky Data Release observed the sky between January and August 2010, observing the sky 1.2 times with four detectors, operating at 3.4, 4.6, 12, and 22$\mu$m. Hereon we refer to \allwise\, photometric bands at [3.4$\mu$m/4.6$\mu$m/12$\mu$m/22$\mu$m] as [$W1/W2/W3/W4$]. The positions of objects in the \wise\, catalog were calibrated to the \twomass\, point source catalog.

The 3-band cryogenic data release contains data from $W1$, 2, and 3, and surveyed $30\%$ of the sky between August and October 2010. During the 3-band cryogenic survey, $W1$ and $W2$ operated with nearly the same sensitivity as during the full survey. Warming of the telescope reduced sensitivity in $W3$ and fully saturated $W4$. The NEOWISE post-cryogenic data release contains $W1$ and $W2$ measurements, with sensitivities close to those obtained during the full cryogenic phase. During this phase, \wise\, surveyed $70\%$ of the sky. Data products from the post-cryogenic release included updated instrumental, astrometric, and photometric calibrations and reduction algorithms, resulting in much lower SNR. The overall number of sources compiled into \allwise\, totals over 747.6 million.

%\subsection{AGB stars from SIMBAD database}
%The SIMBAD database\footnote{\url{http://simbad.u-strasbg.fr/simbad/}} is comprised of astrometry, classifications, and photometry for all types of astronomical objects from a variety of surveys.  Classifications are hierarchical, and are classified with increasing specificity based on the available data, though each star can have multiple classifications. AGB stars have subcategories of C-stars ([C/O] $>$ 1) and S-stars ([C/O] $\sim$ 1). Where subclassification is unclear, the main classification remains as AGB star. OH/IR stars are subsets of peculiar stars that specifically show strong OH maser emission and strong IR emission, identifying them clearly as AGB stars.  Mira-type variables are a subset of variable stars, and are classified based upon their periodicity. Semi-regular variables are not included as they contain an unknown mixture of AGB stars and red supergiants. The sample of AGB stars from SIMBAD was obtained by querying all objects with types classified as C-stars (23,628), S-stars (1,514), AGB stars (4,459), OH/IR stars (1,265), and Mira-type variables (10,828). Without duplicate objects between the five types, there is a total of 39,348 AGB stars from SIMBAD.

%\subsection{AGB stars from optical time-domain surveys}
%Many of the identifications in the SIMBAD catalog are from spectroscopic studies, however many more AGB stars have been found through their characteristic optical periodicity. The \emph{Optical Gravitational Lens Experiment} (OGLE) and the \emph{MAssive Compacy Halo Objects} (MACHO) survey are two time-domain microlensing surveys that have captured and catalogued such variability. In Sections~\ref{sec:macho} and \ref{sec:ogle} we disc.uss the origins of the data from each source and how they were then used for this study

%\subsubsection{MACHO}\label{sec:macho}
%The \emph{MACHO} project \citep{1992ASPC...34..193A,1997ApJ...482...89A} was a two-color optical microlensing survey searching for massive compact halo objects in the Milky Way bulge and the Magellanic Clouds. As a consequence of the time-domain survey, the \emph{MACHO} catalog contains 8 years of observations for several million stars, many of which were observed to be variable. 
%Using Fourier decomposition, \cite{2008PhDT.........6F} identified and classified a population of 46,832 LPVs in the LMC, 3,943 in the SMC {\color{red}[check this]} , and {\color{red}[some other number]} in the Milky Way Bulge with matches to 2MASS. We note that due to high scatter and uncertainty of Galactic objects in period-magnitude space, we only take sequences 1 and 2 (Mira-type variables) for the Milky Way Bulge sample. Each object carries observations in Kron-Cousins $V$ \& $R$, and 2MASS photometry. AGB stars are separated from RGBs by identifying the tip of the RGB ($K_s < 12.3$ in the LMC and $K_s < 12.8$ in the SMC). {\color{red}[some small tidbit about the reduction procedure]}.

\subsection{AGB stars from the OGLE-III Catalog of Variable Stars}\label{sec:ogle}
The \emph{OGLE-III} Catalog of Variable Stars (CVS) \citep{2008AcA....58...69U,2009AcA....59..239S,2011AcA....61..217S} is a subset of the overall \emph{OGLE-III} experiment, containing roughly 10 years of observations in the $V$- and $I$-bands of over 120,000 variable stars. With high-precision photometry \citep[$~0.001$ mag][]{2007AcA....57..201S}, these observations saturate at $I = 12.5$ mag, and are limited at the faint end at $I = 20.5$ mag \citep{2001AcA....51..303Z}.

@@

\subsection{Extragalactic catalogs from SDSS}
SDSS is a multi-year survey of roughly 25\% of the sky collecting both photometric (\emph{u, g, r, i, z}-bands for 700+ million objects) and spectroscopic measurements (1.6+ million) of stars and extragalactic sources \citep{2000AJ....120.1579Y}. Though its ground-based nature and sensitivity to interstellar extinction prevent SDSS from being an all-sky survey, it represents the largest spectrophotometric catalog to date. For this study we collect extragalactic objects from the NYU Value-added Galaxy Catalog, containing objects matched between SDSS and several auxiliary surveys spanning multiple wavelength regimes, and the unofficial SDSS Luminous Red Galaxy sample, isolating high-confidence luminous red galaxies from the SDSS database.

\subsubsection{The NYU Value-added Galaxy Catalog}
In order to cull the full set of SDSS spectrophotometry and produce an easily-referenced extragalactic catalog for investigating galaxy formation and evolution, the NYU Value-added Galaxy Catalog \citep[NYU-VAGC;][]{2005AJ....129.2562B} was created as of SDSS data-release 7 (DR7).

The NYU-VAGC contains matches between SDSS spectrophotometry and the following catalogs: the FIRST radio survey, the 2MASS Point Source Catalog, the 2MASS Extended Source Catalog, the Two-degree Field Galaxy Redshift Survey, the IRAS Point Source Catalog Redshift Survey, and Reference Catalog 3 (RC3.9b). As a consequence, the NYU-VAGC retains SDSS optical photometry in addition to spectral classifications (QSO, galaxy, star) and subclassifications (AGN, starforming galaxy, M3 star, etc.) for 2,506,741 objects across 10,417 square degrees of the SDSS footprint. Main and secondary classifications are made by comparing measured optical spectra to the SDSS spectral library. See \cite{2005AJ....129.2562B} for the full set of reduction and inclusion criteria. This catalog is available via an online data repository\footnote{\url{http://sdss.physics.nyu.edu/vagc/}}, along with a full description of the data and other subsamples not used in this study.

\subsubsection{The Unofficial Luminous Red Galaxy Sample}
Subselected for the study of Baryon Acoustic Oscillations by \cite{2010ApJ...710.1444K}, the Luminous Red Galaxy sample \citep[LRGs;][]{2001AJ....122.2267E} is also sourced from SDSS DR7. LRGs contaminate the color-color space of AGB stars due to the severe redshift of their intrinsically-bright optical fluxes. We obtain the unofficial LRG sample from the data repository\footnote{\url{http://cosmo.nyu.edu/~eak306/SDSS-LRG.html}} of \cite{2010ApJ...710.1444K}. The initial sample is volume-limited, containing 105,631 objects spanning a redshift range of $0.16 < z < 0.47$.

\subsection{WISE+2MASS Young Stellar Objects}
Young Stellar Objects (YSOs) represent a unique contaminant in our search for dust-enshrouded AGB stars. Similar to AGBs, YSOs are luminous sources surrounded by warm, dusty environments. That warm dust can then glow brightly in the MIR, easily contaminating AGB color-color space. To characterize and later eliminate this potential contaminant we select as our base sample 290 YSOs from \cite{2011ApJS..196....4R}, a survey searching for YSOs in the Taurus Molecular Cloud using data from WISE and ancillary data from SDSS and 2MASS. As YSO tend to be embedded within high-extinction clouds of this sort, we expect that the dust-reddened NIR/MIR emission from these objects should match to similar objects in the LMC when we later calibrate our color-color criteria.

### 2015-07-28 paper1/paper1_intro.tex
Edited sections 2.2, 2.3, and 2.5
Changes made up to and including section 2.4

The large distance limit of about 100 kpc for some halo populations (e.g. RR Lyrae and BHB stars 
with SDSS, XXX Sesar 2010 ref and red giants with 2MASS, XXX Majewski ref), as well as about 10 kpc 
for very numerous main sequence stars, is not reached when studying disk component at low galactic
latitudes. As shown, for example, by Berry et al., the extinction due to interstellar dust limits 
the SDSS sample of main sequence stars to heliocentric distances of only a few kpc. In order
to avoid detrimental effect of dust extinction, stellar samples need to be probed at longer infrared
wavelengths. Several infrared surveys covering the Galactic plane have recently become available
(gives refs for WISE and GLIMPSE, VVV: arXiv:1111.5511).

@@

The \emph{Wide-field Infrared Survey Explorer} \citep[\wise, ][]{2010AJ....140.1868W, 2012wise.rept....1C}, is a space observatory that has imaged essentially the entire sky in the MIR (3.4, 4.6, 12, and 22 $\mu$m). \wise\, catalog has been positionally matched to the \twomass\ catalog, with the matched catalog listing NIR and MIR 7-band photometry for
hundreds of millions of sources. Given the depths of the two surveys, this catalog should contain AGB stars to many
kpc beyond the Galactic center. Here we develop selection methods for \agb\ stars using WISE and 2MASS data,
and analyse the resulting samples.

In Section~\ref{sec:data}, we describe in detail the WISE, 2MASS and other auxiliary data used in our study. 
In Section~\ref{sec:criteria}, we use samples of known Galactic and Magellanic \agb\, stars to derive 
WISE-2MASS color-based selection criteria, and calibrate color-absolute magnitude relations. 
In Section~\ref{sec:distribution}, we describe the spatial density distribution of selected \agb\, candidates
from the Milky Way. Our conclusions are summarized in Section~\ref{sec:conclusions}.

>>>>

The large distance limit of about 100 kpc for some halo populations (e.g. RR Lyrae and BHB stars 
with SDSS, \citealt{2010ApJ...708..717S} and red giants with 2MASS, \citealt{2003ApJ...599.1082M}), as well as about 10 kpc 
for very numerous main sequence stars, is not reached when studying disk component at low galactic
latitudes. As shown, for example, by Berry et al., the extinction due to interstellar dust limits 
the SDSS sample of main sequence stars to heliocentric distances of only a few kpc. In order
to avoid detrimental effect of dust extinction, stellar samples need to be probed at longer infrared
wavelengths. Several infrared surveys covering the Galactic plane have recently become available (WISE, \citealt{2010AJ....140.1868W, 2012wise.rept....1C}, GLIMPSE \citealt{2009PASP..121..213C, 2003PASP..115..953B}, VVV \citealt{2012A&A...537A.107S}).

@@

The \emph{Wide-field Infrared Survey Explorer}, \wise, is a space observatory that has imaged essentially the entire sky in the MIR (3.4, 4.6, 12, and 22 $\mu$m). The \wise\, catalog has been positionally matched to the \twomass\ catalog, with the matched catalog listing NIR and MIR 7-band photometry for
hundreds of millions of sources. Given the depths of the two surveys, this catalog should contain AGB stars to many
kpc beyond the Galactic center. Here we develop selection methods for \agb\ stars using WISE and 2MASS data,
and analyse the resulting samples.

In Section~\ref{sec:data}, we describe in detail the WISE, 2MASS and other auxiliary data used in our study and the data reduction process. 
In Section~\ref{sec:criteria}, we use samples of known Galactic and Magellanic \agb\, stars to derive 
WISE-2MASS color-based selection criteria, and calibrate color-absolute magnitude relations. 
In Section~\ref{sec:distribution}, we describe the spatial density distribution of selected \agb\, candidates
from the Milky Way. Our conclusions are summarized in Section~\ref{sec:conclusions}.
